---
title: "BayesMP tutorial"
author: "Zhiguang Huo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(stringsAsFactors = F)

```


## Introduction

A tutorial to guide through the usage for the BayesMP R package.
A real data example of the mltiple tissue mouse metabolism data is used.
The following parts are included:

- How to prepare the input for BayesMP.
- How to perform MCMC.
- How to perform Bayesian Hypothesis testing for meta-analysis.
- How to detect differential expression patterns (metaPattern)

## How to install the package

The package is available on GitHub page ([https://github.com/Caleb-Huo/BayesMP](https://github.com/Caleb-Huo/BayesMP))

To install the package, 

```{r, eval=F}
library(devtools)
install_github("Caleb-Huo/BayesMP")
```

## How to prepare the input for BayesMP.

Note that you may need internet connection to read in the data.

#### How to prepare the input for BayesMP (0), include packages.
```{r}
## include necessary packages
library(BayesMP) ## Include the BayesMP package
library(limma) ## Will perform differential expression analysis
```

#### How to prepare the input for BayesMP (i), read in data.
```{r}
## read in the data

data_brown <- read.csv("https://bayesmp.github.io/data/mouseMetabolism/data_brown.csv", row.names = 1)
data_heart <- read.csv("https://bayesmp.github.io/data/mouseMetabolism/data_heart.csv", row.names = 1)
data_liver <- read.csv("https://bayesmp.github.io/data/mouseMetabolism/data_liver.csv", row.names = 1)

## Verify gene names match across three tissues
all(rownames(data_brown) == rownames(data_heart))
all(rownames(data_brown) == rownames(data_liver))

## Combime these three studies as list
dataExp <- list(brown=data_brown, heart=data_heart, liver=data_liver)

## Check the dimension of the three studies
sapply(dataExp, dim)

## Check the head of the three studies
sapply(dataExp, head)

## perform differential expression analysis for each of these three tissues.

## Create an empty matrix to store Z value. 
## Each row represents a gene and each column represent a study/tissue. 
## Note that Z value matrix is the input for BayesMP method. 
## Z value can be calculated by inverse CDF of the p-value from differential expression analysis. 
## A positive Z value indicates that the gene is upregulated in that study/tissue.

Z <- matrix(0,nrow=nrow(dataExp[[1]]),ncol=length(dataExp))
rownames(Z) <- rownames(dataExp[[1]])
colnames(Z) <- names(dataExp)
```


#### How to prepare the input for BayesMP (ii), perform differential expression analysis in each study
```{r}
for(s in 1:length(dataExp)){
	adata <- dataExp[[s]]
	ControlLabel = grep('wt',colnames(adata))
	caseLabel = grep('LCAD',colnames(adata))
	label <- rep(NA, ncol(adata))
	label[ControlLabel] = 0
	label[caseLabel] = 1
	
	design = model.matrix(~label)	## design matrix
	fit <- lmFit(adata,design)		## fit limma model
	fit <- eBayes(fit)		
	
	aeffectsize <- fit$coefficients[,2]	## get effect sizes
	Z[aeffectsize>0,s] <- -qnorm(fit$p.value[aeffectsize>0,2]/2)
	Z[aeffectsize<0,s] <- qnorm(fit$p.value[aeffectsize<0,2]/2)
	
}
```

#### How to prepare the input for BayesMP (iii), check the head of the input matrix Z.

```{r}
## checkout top rows of Z
head(Z)
```

## BayesMP modeling via MCMC.
There are two models.

- BayesMP_DP will model the alternative distribution via Dirichlet process.
- BayesMP_EP will model the alternative distribution via Gaussian distribution.

Theoritically BayesMP_DP will generate better results since the Dirichlet process is a non-parametric model and robust again model perturbations. 
But in reality, BayesMP_EP will also acheive accurate results with much faster speed.



```{r}
## WD <- '~/Desktop/'
## setwd(WD) ## You can set the working directory here. Some MCMC results will be saved here.

niter <- 1000 ## number of iterations.
burnin <- 200 ## number of burnin samples. Note the burnin samples will be disgarded.

set.seed(15213)	## set random seed

## writeHSall=T will save the intermediate results for the Bayesian hypothesis testing settings.
## writeY=T will save the intermediate results for the posterior samples of Y, which is used as the input for metaPattern detection.
## If you want to track the MCMC samples for other paramters including gamma, Pi, Delta, 
## One can save these results by set writeGamma, writePi, writeDelta to be TRUE.
## One can set gamma to be fixed by setting updateGamma = TRUE, if he has a good estimate of initial gamma.

system.time(BayesMP_DP(Z,niter=niter, burnin=burnin, writeY = T, writeHSall=T))
```

## Task I, meta analysis. There are three hypothesis testing settings.

- Settings $HS_B$ targets biomarkers that are DE in one or more studies:
    - $\Omega_{B}$: $\Omega_{B}^1 =\{\vec{\theta}_g:  \sum_{s=1}^S\mathbb{I}(\theta_{gs}\ne 0) \ge 1 \}$.
- Settings $HS_A$ targets biomarkers that are DE in all studies:
    - $\Omega_{\bar{A}}$: $\Omega_{\bar{A}}^1 =\{\vec{\theta}_g:  \sum_{s=1}^S\mathbb{I}(\theta_{gs}\ne 0)=S \}$.
- Settings $HS_r$ targets biomarkers that are DE in at least $r$ studies:
    - $\Omega_{\bar{r}}$: $\Omega_{\bar{r}}^1 =\{\vec{\theta}_g:  \sum_{s=1}^S\mathbb{I}(\theta_{gs}\ne 0)\ge r \}$.


```{r}
HSallRes <- read.table('BayesMP_DP_HSall.txt') ## read in the intermediate results for the Bayesian hypothesis testing
 
## Omega(B)
HSb_belief <- HSallRes[,1]/(niter - burnin) ## Bayesian belief for Omega(B)
HSb_qvalue <- BayesianFDR(HSb_belief) ## Bayesian FDR for each gene
sum(HSb_qvalue<0.05)
sum(HSb_qvalue<0.01)


## Omega(A)
S <- ncol(Z)
HSa_belief <- HSallRes[,S]/(niter - burnin) ## Bayesian belief for Omega(A)
HSa_qvalue <- BayesianFDR(HSa_belief) ## Bayesian FDR for each gene
sum(HSa_qvalue<0.05)

## Omega(r)
r <- 2
HSr_belief <- HSallRes[,r]/(niter - burnin) ## Bayesian belief for Omega(r)
HSr_qvalue <- BayesianFDR(HSr_belief) ## Bayesian FDR for each gene
sum(HSr_qvalue<0.05)
 
```

## Task II, differential expression meta-pattern (MetaPattern)
#### Task II (i), read in MCMC posterior samples
```{r}
con  <- file('BayesMP_DP_Y.txt', open = "r")

G <- nrow(Z)
S <- ncol(Z)

## the resYplus and resYminus matrices are used to save "latent expression"
resYplus <- matrix(0,G,S)
resYminus <- matrix(0,G,S)


i = 1
while (length(oneLine <- readLines(con, n = 1, warn = FALSE)) > 0) {
  if(i>burnin){
	  #print(i)
	  seven = strsplit(oneLine, "\t")[[1]]
	  thisY <- matrix(as.numeric(seven),G,S)
  	
	  ## for individual studies
	  resYplus[thisY>0] <- resYplus[thisY>0] + 1
	  resYminus[thisY<0] <- resYminus[thisY<0] + 1
  }    
  i = i + 1
} 

close(con)
```

#### Task II (ii), detect meta pattern. We use the tight clustering algorithm to detect metaPatterns. In fact, one can also use other clustering algorithm to detect gene clusters.
```{r}
## we only consider HSb_qvalue<0.01 for the meta-pattern detection
resYplus_DE <- resYplus[HSb_qvalue<0.01,]
resYminus_DE <- resYminus[HSb_qvalue<0.01,]

## tight clustering
dissimilarity <- distance(resYplus_DE, resYminus_DE, niter - burnin) ## calculate dissimilarity matrix for each pair of genes.
atarget <- 2

## Alternatively, one may try the consensus clustering algorithm to detect gene modules.
## perform the tight clustering to identify meta-pattern modules. 
## k.min controls the size of the final meta module. 
## A small k.min will result in large module size and a large k.min will result in small module size.
tightClustResult <- tightClustPam(dissimilarity, target=atarget, k.min=20) 
```


#### Task II (iii), visualization
```{r, fig.show='hold'}
## visualize the posterior probablity

for(i in 1:atarget){
	clustSelection <- tightClustResult$cluster == i
	numS = ncol(resYplus_DE)
	barData <- cbind(resYplus_DE,resYminus_DE)[clustSelection,]/(niter - burnin)
	barMean <- colMeans(barData)
	barSd <- apply(barData,2,sd)

	mp <- barplot(barMean, axes=FALSE, axisnames=FALSE, ylim=c(0, 1.2),
	              col=c(rep('red',numS), rep('blue',numS)), main=paste('target',i,'\n n =',sum(clustSelection)), 
				  xlab="Study", ylab="prosterior probablity")
	axis(1, labels=c(paste(1:numS,"+",sep=''), paste(1:numS,"-",sep='')), at = mp)
	axis(2, at=seq(0 , 1, by=0.2))
	 
	box()

	# Plot the vertical lines of the error bars
	# The vertical bars are plotted at the midpoints
	segments(mp, barMean - barSd, mp, barMean + barSd, lwd=2)
	# Now plot the horizontal bounds for the error bars
	# 1. The lower bar
	segments(mp - 0.1, barMean - barSd, mp + 0.1, barMean - barSd, lwd=2)
	# 2. The upper bar
	segments(mp - 0.1, barMean + barSd, mp + 0.1, barMean + barSd, lwd=2)
}
```

```{r, fig.show='hold'}
## visualize the heatmap of the first metaPattern module.

for(s in 1:length(dataExp)){
	adata <- dataExp[[s]]
	aname <- names(dataExp)[s]
	bdata <- adata[HSb_qvalue<0.01, ][tightClustResult$cluster == 1 ,]
	cdata <- as.matrix(bdata)
	ddata <- t(scale(t(cdata))) ## standardize the data such that for each gene, the mean is 0 and sd is 1.

	ColSideColors <- rep("black", ncol(adata))
	ColSideColors[grep('LCAD',colnames(adata))] <- "red"
	
	B <- 16
  redGreenColor <- rgb(c(rep(0, B), (0:B)/B), c((B:0)/16, rep(0, B)), rep(0, 2*B+1))
	heatmap(ddata,Rowv=NA,ColSideColors=ColSideColors,col= redGreenColor ,scale='none',Colv=NA, main=aname)
}
```


